PhD Thesis

### 2016-12-02 pldi14/sec-practice.tex
Prepare integration: fix conflicts with polytable

\pg{Why does 'target language' show up all of a sudden? Shouldn't
  we say that we generate Scala code for execution
  \textbf{elsewhere}? Right now, we say it in next section. Confusing.}
%
\Cref{fig:primitives} shows a simplified
Scala implementation of the primitives
used in \cref{fig:case-study-pseudocode}.
As bag primitives, we provide constructors and a fold operation,
following \citet{GlucheGrust97Incr}. The constructors for bags
are $\Empty$ (constructing the empty bag), $\SINGLETON$
(constructing a bag with one element), $\MERGE$ (constructing the
merge of two bags) and $\NEGATE$ ($\Negate{b}$ constructs a bag
with the same elements as $b$ but negated multiplicities); all but $\SINGLETON$ represent
abelian group operations.
%
Unlike for usual ADT constructors, the same bag can be
constructed in different ways, which are equivalent by the equations defining abelian groups;
for instance, since $\MERGE$ is commutative, $\Merge{x}{y} = \Merge{y}{x}$.
%
Folding on a bag will represent the bag through constructors in
an arbitrary way, and then replace constructors with arguments;
to ensure a well-defined result, the arguments of fold should
respect the same equations, that is, they should form an abelian group;
for instance, the binary operator should be commutative.
%
Hence, the fold operator $\FOLDBAG$ can be defined to take a
function (corresponding to $\SINGLETON$) and an abelian group
(for the other constructors). $\FOLDBAG$ is then defined by equations:
%
\begin{alignat*}{2}
&  \HasType{\FOLDBAG}{\mathbf{Group}\; \Gt \to (\Gs \to \Gt) \to &&\Bag[\Gs] \to \Gt}\\
&  \App{\FoldBag {g @ (\_, \boxplus, \boxminus, e)}{f}}{\Empty}
    &&= e \displaybreak[0]\\
&  \App{\FoldBag {g @ (\_, \boxplus, \boxminus, e)}{f}}{\Merge* {b_1} {b_2}}
   &&= \App{\FoldBag{g}{f}}{b_1} \displaybreak[0]\\
&&& \boxplus
  \App{\FoldBag{g}{f}}{b_1} \displaybreak[0]\\
&  \App{\FoldBag {g @ (\_, \boxplus, \boxminus, e)}{f}}{\Negate* b}
   && = \boxminus \; \App*{\FoldBag{g}{f}}{b}\displaybreak[0]\\
&  \App{\FoldBag {g @ (\_, \boxplus, \boxminus, e)}{f}}{\Singleton*{v}}
   &&= \App{f}{v}
\end{alignat*}
%
If $g$ is a group,
these equations specify $\App{\FOLDBAG} g$ precisely~\citep{GlucheGrust97Incr}.
%
Moreover, the first three equations mean that $\FoldBag{g}{f}$ is the \emph{abelian group
  homomorphism} between the abelian group on bags and the group
$g$ (because those equations coincide with the definition).
%
\Cref{fig:primitives} shows an implementation of $\FOLDBAG$ as
specified above.
Moreover, all functions which deconstruct a bag can be expressed in
terms of $\FOLDBAG$ with suitable arguments.
%
For instance, we can sum the elements of a bag of integers with
$\FoldBag{\Term{gZ}}{\Lam*{x}{x}}$, where
$\Term{gZ}$ is the abelian group on integers defined in
\cref{ssec:change-structures}.
Users of $\FOLDBAG$ can define different abelian groups to specify
different operations (for instance, to multiply floating-point numbers).

@@

These are the primitives on maps.
\begin{align*}
\HasTypeAligned\MAKEMAP
  {\Fun\Gk{\Fun\Gi{\HashMap\Gk\Gi}}}
\\
\HasTypeAligned\LOOKUP
  {\Fun\Gk{\Fun{\Map\Gk\Gi}{\Maybe\Gi}}}
\\
\HasTypeAligned\LIFTGROUP
  \Fun{\Abelian\Gi}{\Abelian{\HashMap\Gk\Gi}}
\\
\HasTypeAligned\FOLDMAP
  \Fun{\Abelian\Gi}
    {\Fun{\Abelian\Gt} \\ & \qquad
      {\Fun{ \Fun*\Gk{\Fun\Gi\Gt} }
        {\Fun{\HashMap\Gk\Gi} \Gt}}}
\end{align*}

\begin{definition}[Semantics of map primitives]~
\begin{subdefinition}
\item $\Eval{\Map\Gk\Gi}$ is the set of partial functions from
$\Eval\Gk$ to $\Eval\Gi$ defined only at a finite number of
points.
\item $\MAKEMAP$ evaluates to the function that converts a
key-value pair to the corresponding partial function defined at
1 point.
\item $\LOOKUP$ evaluates to the standard lookup function on
maps.
\item $\LIFTGROUP$ evaluates to a function that maps abelian
groups $G_\Gi$ over $\Eval\Gi$ to the abelian group formed by
direct product of copies of $G_\Gi$ indexed by $\Eval\Gk$. Its
identity element, binary operation and inverse function are the
empty map, map merge and map negation.
\item Given a map $\HasType m {\HashMap\Gk\Gi}$, abelian groups
$G_\Gi$, $G_\Gt$, and a function $\HasType f
{\Fun\Gk{\Fun\Gi\Gt}}$ such that $\App*fk$ is a homomorphism from
$G_\Gi$ to $G_\Gt$ for every $\HasType k\Gk$, the term
\[
\App{\App{\App{\App\FOLDMAP{G_\Gi}}{G_\Gt}}f}m
\]
evaluates to a value in $\Eval\Gt$ by the following steps (we
write $m(k)$ for the value of the map $m$ at key $k$).
\begin{enumerate}
\item Replace each $m(k)$ by $\App*{\App fk}{m(k)}$ to create the
map $\HasType{m'}{\HashMap\Gk\Gt}$.
\item Sum all values of $m'$ by the binary operation of $G_\Gt$.
\end{enumerate}
Since both steps are homomorphisms,
$\App*{\App{\App\FOLDMAP{G_\Gi}}{G_\Gt}}f$ is a homomorphism
from $\App*\LIFTGROUP{G_\Gi}$ to $G_\Gt$.
% to the unique homomorphism%
% \footnote{The existence and uniqueness of such a homomorphism is
% guaranteed by the universal property of indexed direct product
% as a product in the category of abelian groups.} %
% from $\App*\LIFTGROUP{G_\Gi}$ to $G_\Gt$ that agrees with
% $\App*fk$ for every $\HasType k \Gk$.
%
\end{subdefinition}
\end{definition}

@@

% OLDSEC: abelian groups
% Refer to it when discussing AbelianDerivation traits.
\begin{oldSec}
Our approach is parametric in the concrete representation of
changes of type $\Change{\tau}$ as long as changes support the
operations listed in \cref{fig:change-operations}).
The update operator $\Apply{\D x}{x}$ updates a value $x$ with a
change $\D x$. The difference operator $\Diff{y}{x}$ computes the
change from $x$ to $y$.
And the nil change operator $\Nil{x}$
denotes the change that updates the value $x$ into itself.

@@

Our presentation of the behavior of changes on a datatype forms
an algebraic specification (excluding the concept of validity and
the properties which only apply to valid changes). In particular,
we can define a signature having sorts $\tau$ and $\Change{\tau}$,
the functions from \cref{fig:change-operations}
and equations (or axioms):
\begin{align*}
\Apply{\Diff*{s₂}{s₁}}{s₁} &= s₂\\
\Apply{\Nil{s}}{s} &= s\\
\end{align*}
\pg{Check which ones we have already proven.}

>>>>

\pg{Why does 'target language' show up all of a sudden? Shouldn't
  we say that we generate Scala code for execution
  \textbf{elsewhere}? Right now, we say it in next section. Confusing.}
%
\Cref{fig:primitives} shows a simplified
Scala implementation of the primitives
used in \cref{fig:case-study-pseudocode}.
As bag primitives, we provide constructors and a fold operation,
following \citet{GlucheGrust97Incr}. The constructors for bags
are $\Empty$ (constructing the empty bag), $\SINGLETON$
(constructing a bag with one element), $\MERGE$ (constructing the
merge of two bags) and $\NEGATE$ ($\Negate{b}$ constructs a bag
with the same elements as $b$ but negated multiplicities); all but $\SINGLETON$ represent
abelian group operations.
%
Unlike for usual ADT constructors, the same bag can be
constructed in different ways, which are equivalent by the equations defining abelian groups;
for instance, since $\MERGE$ is commutative, $\Merge{x}{y} = \Merge{y}{x}$.
%
Folding on a bag will represent the bag through constructors in
an arbitrary way, and then replace constructors with arguments;
to ensure a well-defined result, the arguments of fold should
respect the same equations, that is, they should form an abelian group;
for instance, the binary operator should be commutative.
%
Hence, the fold operator $\FOLDBAG$ can be defined to take a
function (corresponding to $\SINGLETON$) and an abelian group
(for the other constructors). $\FOLDBAG$ is then defined by equations:
%
\begin{alignat*}{2}
&  \HasType{\FOLDBAG}{\mathbf{Group}\; \Gt \to (\Gs \to \Gt) \to &&\Bag[\Gs] \to \Gt}\\
&  \App{\FoldBag {g @ (\_, \boxplus, \boxminus, e)}{f}}{\Empty}
    &&= e \displaybreak[0]\\
&  \App{\FoldBag {g @ (\_, \boxplus, \boxminus, e)}{f}}{\Merge* {b_1} {b_2}}
   &&= \App{\FoldBag{g}{f}}{b_1} \displaybreak[0]\\
&&& \boxplus
  \App{\FoldBag{g}{f}}{b_1} \displaybreak[0]\\
&  \App{\FoldBag {g @ (\_, \boxplus, \boxminus, e)}{f}}{\Negate* b}
   && = \boxminus \; \App*{\FoldBag{g}{f}}{b}\displaybreak[0]\\
&  \App{\FoldBag {g @ (\_, \boxplus, \boxminus, e)}{f}}{\SingletonT*{v}}
   &&= \App{f}{v}
\end{alignat*}
%
If $g$ is a group,
these equations specify $\App{\FOLDBAG} g$ precisely~\citep{GlucheGrust97Incr}.
%
Moreover, the first three equations mean that $\FoldBag{g}{f}$ is the \emph{abelian group
  homomorphism} between the abelian group on bags and the group
$g$ (because those equations coincide with the definition).
%
\Cref{fig:primitives} shows an implementation of $\FOLDBAG$ as
specified above.
Moreover, all functions which deconstruct a bag can be expressed in
terms of $\FOLDBAG$ with suitable arguments.
%
For instance, we can sum the elements of a bag of integers with
$\FoldBag{\Term{gZ}}{\Lam*{x}{x}}$, where
$\Term{gZ}$ is the abelian group on integers defined in
\cref{ssec:change-structures}.
Users of $\FOLDBAG$ can define different abelian groups to specify
different operations (for instance, to multiply floating-point numbers).

@@

These are the primitives on maps.
\begin{align*}
\HasTypeAligned\MAKEMAP
  {\Fun\Gk{\Fun\Gi{\HashMap\Gk\Gi}}}
\\
\HasTypeAligned\LOOKUP
  {\Fun\Gk{\Fun{\MapT\Gk\Gi}{\Maybe\Gi}}}
\\
\HasTypeAligned\LIFTGROUP
  \Fun{\Abelian\Gi}{\Abelian{\HashMap\Gk\Gi}}
\\
\HasTypeAligned\FOLDMAP
  \Fun{\Abelian\Gi}
    {\Fun{\Abelian\Gt} \\ & \qquad
      {\Fun{ \Fun*\Gk{\Fun\Gi\Gt} }
        {\Fun{\HashMap\Gk\Gi} \Gt}}}
\end{align*}

\begin{definition}[Semantics of map primitives]~
\begin{subdefinition}
\item $\Eval{\MapT\Gk\Gi}$ is the set of partial functions from
$\Eval\Gk$ to $\Eval\Gi$ defined only at a finite number of
points.
\item $\MAKEMAP$ evaluates to the function that converts a
key-value pair to the corresponding partial function defined at
1 point.
\item $\LOOKUP$ evaluates to the standard lookup function on
maps.
\item $\LIFTGROUP$ evaluates to a function that maps abelian
groups $G_\Gi$ over $\Eval\Gi$ to the abelian group formed by
direct product of copies of $G_\Gi$ indexed by $\Eval\Gk$. Its
identity element, binary operation and inverse function are the
empty map, map merge and map negation.
\item Given a map $\HasType m {\HashMap\Gk\Gi}$, abelian groups
$G_\Gi$, $G_\Gt$, and a function $\HasType f
{\Fun\Gk{\Fun\Gi\Gt}}$ such that $\App*fk$ is a homomorphism from
$G_\Gi$ to $G_\Gt$ for every $\HasType k\Gk$, the term
\[
\App{\App{\App{\App\FOLDMAP{G_\Gi}}{G_\Gt}}f}m
\]
evaluates to a value in $\Eval\Gt$ by the following steps (we
write $m(k)$ for the value of the map $m$ at key $k$).
\begin{enumerate}
\item Replace each $m(k)$ by $\App*{\App fk}{m(k)}$ to create the
map $\HasType{m'}{\HashMap\Gk\Gt}$.
\item Sum all values of $m'$ by the binary operation of $G_\Gt$.
\end{enumerate}
Since both steps are homomorphisms,
$\App*{\App{\App\FOLDMAP{G_\Gi}}{G_\Gt}}f$ is a homomorphism
from $\App*\LIFTGROUP{G_\Gi}$ to $G_\Gt$.
% to the unique homomorphism%
% \footnote{The existence and uniqueness of such a homomorphism is
% guaranteed by the universal property of indexed direct product
% as a product in the category of abelian groups.} %
% from $\App*\LIFTGROUP{G_\Gi}$ to $G_\Gt$ that agrees with
% $\App*fk$ for every $\HasType k \Gk$.
%
\end{subdefinition}
\end{definition}

@@

% OLDSEC: abelian groups
% Refer to it when discussing AbelianDerivation traits.
\begin{oldSec}
Our approach is parametric in the concrete representation of
changes of type $\Change{\tau}$ as long as changes support the
operations listed in \cref{fig:change-operations}).
The update operator $\Apply{\D x}{x}$ updates a value $x$ with a
change $\D x$. The difference operator $\Diff{y}{x}$ computes the
change from $x$ to $y$.
And the nil change operator $\NilC{x}$
denotes the change that updates the value $x$ into itself.

@@

Our presentation of the behavior of changes on a datatype forms
an algebraic specification (excluding the concept of validity and
the properties which only apply to valid changes). In particular,
we can define a signature having sorts $\tau$ and $\Change{\tau}$,
the functions from \cref{fig:change-operations}
and equations (or axioms):
\begin{align*}
\Apply{\Diff*{s₂}{s₁}}{s₁} &= s₂\\
\Apply{\NilC{s}}{s} &= s\\
\end{align*}
\pg{Check which ones we have already proven.}
